First Order Alternation

### 2018-06-27 quantifiers/draft.tex
ok

For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. To simplify the
presentation, we consider two sorts $\Data$ and $\Bool$, where $\Data$
ranges over an infinite domain and $\Bool$ has the usual
interpretation. The $\Data$ sort is equipped with a finite set of
function symbols $f : \Data^{\#(f)} \rightarrow \Data$, where
$\#(f)\geq0$ denotes the number of arguments (arity) of $f$. When
$\#(f)=0$, we say that $f$ is a constant, and denote by $\top$ and
$\bot$ the only constants of sort $\Bool$, representing the true and
false values, respectively. A \emph{predicate} is a function symbol $p
: \Data^{\#(p)} \rightarrow \Bool$, denoting a relation of arity
$\#(p)$ and we write $\preds$ for the set of predicates. In the
following, we shall consider that the interpretation of all function
symbols $f : \Data^{\#(f)} \rightarrow \Data$ that are not predicates
is fixed and relative to the interpretation of the $\Data$ sort.

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over the interpretation of the $\Data$ sort. Terms
are either constants of sort $\Data$, variables or function
applications. The set of \emph{positive} first-order formulae is
defined by the syntax below:
\[\phi := t \teq s \mid t \not\teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \phi_1 \wedge \phi_2 \mid \phi_1 \vee \phi_2 \mid \exists x ~.~
\phi_1 \mid \forall x ~.~ \phi_1\] where $t,s,t_1,\ldots,t_{\#(p)}$
denote terms. We denote by $\fv{}{\phi}$ the set of free variables in
$\phi$. A \emph{sentence} is a formula $\phi$ in which each variable
occurs under the scope of a quantifier, i.e. $\fv{}{\phi} =
\emptyset$. The size $\len{\phi}$ is the number of symbols occuring in
$\phi$.

@@

An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
obtained by replacing each variable $x$ by the value $\nu(x)$ and
evaluating each function application. For a formula $\phi$, we define
the forcing relation $\I,\nu \models \phi$ recursively on the
structure of $\phi$, as follows:
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models t \not\teq s & \iff & t^\nu \neq s^\nu \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \phi_1 \vee \phi_2 & \iff & \I,\nu \models \phi_i \text{, for some } i=1,2 \\
\I,\nu \models \forall x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for all } d \in \Data \\
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
\end{array}
\]
For a formula $\phi$, we define $\sem{\phi}_\nu \isdef \set{\I \mid
  \I,\nu \models \phi}$, where $\nu(x)$ is defined for each $x \in
\fv{}{\phi}$, and drop the subscript for sentences. A sentence $\phi$
is \emph{satisfiable} if $\sem{\phi}\neq\emptyset$, an element of
$\sem{\phi}$ being called a \emph{model} of $\phi$.

@@

Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
  events}. Given a finite set of variables $X$, we denote by $X
\mapsto \Data^\I$ the set of valuations of the variables $X$ and
$\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly infinite
set of \emph{data symbols}. A \emph{data word} (word) is a finite
sequence $w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n : X
\rightarrow \Data$ are valuations. Given a word $w$, we denote by
$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
$\data{w}$ the valuation associating each time-stamped variable
$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
$\Sigma^*$ the set of finite sequences of input events and by
$\Sigma[X]^*$ the set of data words over the variables $X$.

A \emph{first-order alternating automaton} (automaton) is a tuple $\A
= \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is the finite
event alphabet, $X$ is a finite set of variables, $Q$ is a finite set
of predicates, $\iota$ is a satisfiable sentence, $F \subseteq Q$ is a
set of \emph{final predicates} and $\Delta$ is a set of
\emph{transition rules} of the form \(q(y_1,\ldots,y_{\#(q)})
\arrow{a(X)}{} \psi\), where $q \in Q$ is a predicate, $a \in \Sigma$
is an input event, $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$,
$\psi$ is a positive formula such that $\fv{}{\phi} \subseteq X \cup
\set{y_1,\ldots,y_{\#(q)}}$ and each predicate that occurs in $\psi$
belongs also to $Q$. The \emph{size} of $\A$ is defined as $\len{\A} =
\len{\iota} + \sum_{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}
\len{\psi}$.

Intuitively, $a$ is the input event and $X$ are the input data values
that trigger the transition, whereas $q$ and $y_1,\ldots,y_{\#(q)}$
are the current control state and data values pertaining to that
state, respectively. Without loss of generality, we consider, for each
predicate $q \in Q$ and each input event $a \in \Sigma$, at most one
such rule --- two or more rules can be joined using disjunction.

The execution semantics of of automata is given in close analogy with
the case of boolean alternating automata, with transition rules of the
form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
a positive boolean combination of such constants. For instance, $q_0
\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
choose to transition in either both $q_1$ and $q_2$ or in $q_3$
alone. This intuition leads to saying that the steps of the automaton
are defined by the minimal boolean models of the transition
formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, but not
$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
which is a model but is not minimal\footnote{To avoid the confusion of
  considering potentially non-minimal models, the original definition
  of alternating finite-state automata \cite{ChandraKozenStockmeyer81}
  uses boolean valuations instead of formulae.}.

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
set $\mathcal{S}$ of interpretations, we define $\cube{\mathcal{S}}
\isdef \set{\cube{\I} \mid \I \in \mathcal{S}}$. Given a word
$w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in \Sigma[X]^*$, an
\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is
a (possibly infinite) forest $\F_w = \set{T_1,T_2,\ldots}$, where each
tree $T_i$ is defined below:
\begin{compactitem}
\item $T_i$ is rooted in a configuration from some cube from
  $\cube{\minsem{\iota}}$ --- because $\iota$ is satisfiable, such a cube
  always exists,
%
\item all paths from the root to a leaf of $T_i$ have length $n$, and
%
\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on the level $j \in
  [n-1]$ in $T_i$, then its children form a cube from
  $\cube{\minsem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
  $\eta$ is a valuation that assigns each variable $y_k$ the value
  $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
  $\nu_{j+1}(x)$ --- because $\fv{}{\psi} \subseteq X \cup
  \set{y_1,\ldots,y_{\#(q)}}$, the values assigned by $\eta$ to
  variables outside $X \cup \set{y_1,\ldots,y_{\#(q)}}$ have no impact
  on $\cube{\minsem{\psi}_\eta}$.
\end{compactitem}
A forest $\F_w$ is said to be \emph{accepting} if and only if the
frontier of each tree $T \in \F_w$ is labeled with configurations
$q(d_1,\ldots,d_{\#(q)})$ such that $q \in F$ is a final predicate. In
this case, we say that $\A$ accepts $w$ and denote by $\lang{\A}$ the
set of words accepted by $\A$. In this paper, we answer the following
questions: \begin{compactenum}
\item \emph{closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

In the upcoming developments it is more convenient to work with path
formulae defining executions.  Let $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the rest of
this session. For any $i \in \nat$, we denote by $\stamp{Q}{i} =
\set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} = \set{\stamp{x}{i}
  \mid x \in X}$ the sets of time-stamped predicates and variables,
respectively. For a formula $\psi$ and $i \in \nat$, we define
$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
formula in which all input variables and state predicates are replaced
by their time-stamped counterparts.

Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
\emph{path formula} of $\alpha$ in $\A$ is defined as follows: 
\[\pathform{\A}{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow \stamp{\psi}{i}\]
Observe that path formulae are not positive, due to the occurrence of
predicate atoms to the left of the implication as in
$\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow
\stamp{\psi}{i}$. The intuition is that, whenever the automaton is in
a configuration $q(d_1, \ldots, d_{\#(q)})$ at step $i-1$, it proceeds
at step $i$, according to $\psi$.

To formalize the relation between the low-level configuration-based
execution semantics and the symbolic path formulae, consider a word
$w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any execution forest $\F_w$ of
$\A$ over $w$ is associated a unique interpretation defined as
$\I_{[\F_w]}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}}
  \mid q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in $\F_w$}}$
for each $q \in \preds$ and $i \in [n]$.

\begin{lemma}\label{lemma:path-formula}
  For any word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have: 
  \[\minsem{\pathform{\A}{w_\Sigma}}_{w_\Data} = \set{\I_{[\F_w]} \mid \F_w
    \text{ is an execution of $\A$ over $w$}}\enspace.\]
\end{lemma}

>>>>

For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. To simplify the
presentation, we consider two sorts $\Data$ and $\Bool$, where $\Data$
is an infinite domain and $\Bool$ is the set of truth values $\top$
(true) and $\bot$ (false). The $\Data$ sort is equipped with finitely
many function symbols $f : \Data^{\#(f)} \rightarrow \Data$, where
$\#(f)\geq0$ denotes the number of arguments (arity) of $f$. When
$\#(f)=0$, we say that $f$ is a constant. A \emph{predicate} is a
function symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a
relation of arity $\#(p)$ and we write $\preds$ for the set of
predicates. In the following, we shall consider that the
interpretation of all function symbols $f : \Data^{\#(f)} \rightarrow
\Data$ that are not predicates is fixed and relative to the
interpretation of the $\Data$ sort. For simplicity, we further blur
the notational distinction between function symbols and their
interpretations.

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over the interpretation of the $\Data$ sort. Terms
are either constants of sort $\Data$, variables or function
applications. The set of \emph{positive} first-order formulae is
defined by the syntax below:
\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
$\phi_1 \vee \phi_2$, $\phi_1 \Rightarrow \phi_2$ and $\forall x ~.~
\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
$\fv{}{\phi}$ the set of free variables in $\phi$. A \emph{sentence}
is a formula $\phi$ in which each variable occurs under the scope of a
quantifier, i.e. $\fv{}{\phi} = \emptyset$. A formula is
\emph{positive} if each predicate symbol occurs under an even number
of negations. The size $\len{\phi}$ of a formula $\phi$ is the number
of symbols needed to write it down.

@@

An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
obtained by replacing each variable $x$ by the value $\nu(x)$ and
evaluating each function application. For a formula $\phi$, we define
the forcing relation $\I,\nu \models \phi$ recursively on the
structure of $\phi$, as follows:
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
\end{array}
\]
For a formula $\phi$, we define $\sem{\phi}_\nu \isdef \set{\I \mid
  \I,\nu \models \phi}$, where $\nu(x)$ is defined for each $x \in
\fv{}{\phi}$, and drop the subscript for sentences. A sentence $\phi$
is \emph{satisfiable} if $\sem{\phi}\neq\emptyset$, an element of
$\sem{\phi}$ being called a \emph{model} of $\phi$.

@@

Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
  events}. Given a finite set of variables $X$, we denote by $X
\mapsto \Data$ the set of valuations of the variables $X$ and
$\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly infinite
set of \emph{data symbols} $(a,\nu)$, where $a$ is an input symbol and
$\nu$ is a valuation. A \emph{data word} (simply called word in the
following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
$\data{w}$ the valuation associating each time-stamped variable
$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
$\Sigma^*$ the set of finite sequences of input events and by
$\Sigma[X]^*$ the set of data words over the variables $X$.

A \emph{first-order alternating automaton} (simply called automaton in
the following) is a tuple $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$,
where $\Sigma$ is the finite event alphabet, $X$ is a finite set of
variables, $Q$ is a finite set of predicates, $\iota$ is a satisfiable
sentence, $F \subseteq Q$ is a set of \emph{final predicates} and
$\Delta$ is a set of \emph{transition rules} of the form
\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
predicate, $a \in \Sigma$ is an input event, $X \cap
\set{y_1,\ldots,y_{\#(q)}} = \emptyset$, $\psi$ is a positive formula
such that $\fv{}{\phi} \subseteq X \cup \set{y_1,\ldots,y_{\#(q)}}$
and each predicate that occurs in $\psi$ belongs also to $Q$. The
\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}} \len{\psi}$.

The intuition of a transition rule \(q(y_1,\ldots,y_{\#(q)})
\arrow{a(X)}{} \psi\) is the following: $a$ is the input event and $X$
are the input data values that trigger the transition, whereas $q$ and
$y_1,\ldots,y_{\#(q)}$ are the current control state and data values
in that state, respectively. Without loss of generality, we consider,
for each predicate $q \in Q$ and each input event $a \in \Sigma$, at
most one such rule --- two or more rules can be joined using
disjunction.

The execution semantics of automata is given in close analogy with
the case of boolean alternating automata, with transition rules of the
form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
a positive boolean combination of such constants. For instance, $q_0
\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
choose to transition in either both $q_1$ and $q_2$ or in $q_3$
alone. This intuition leads to saying that the steps of the automaton
are defined by the minimal boolean models of the transition
formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, but not
$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
which is a model but is not minimal\footnote{To avoid the confusion of
  considering potentially non-minimal models, the original definition
  of alternating finite-state automata \cite{ChandraKozenStockmeyer81}
  uses boolean valuations instead of formulae.}.

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
set $\mathcal{S}$ of interpretations, we define $\cube{\mathcal{S}}
\isdef \set{\cube{\I} \mid \I \in \mathcal{S}}$. Given a word
$w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in \Sigma[X]^*$, an
\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is
a (possibly infinite) forest $\F = \set{T_1,T_2,\ldots}$, where each
tree $T_i$ is defined below:
\begin{compactitem}
\item $T_i$ is rooted in a configuration from some cube from
  $\cube{\minsem{\iota}}$. % --- because $\iota$ is satisfiable, such
                           % a cube always exists,
%
\item all paths from the root to a leaf of $T_i$ have length $n$, and
%
\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on the level $j \in
  [n-1]$ in $T_i$, then its children form a cube from
  $\cube{\minsem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
  $\eta$ is a valuation that assigns each variable $y_k$ the value
  $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
  $\nu_{j+1}(x)$. % --- because $\fv{}{\psi} \subseteq X \cup
                  % \set{y_1,\ldots,y_{\#(q)}}$, the values assigned
                  % by $\eta$ to variables outside $X \cup
                  % \set{y_1,\ldots,y_{\#(q)}}$ have no impact on
                  % $\cube{\minsem{\psi}_\eta}$.
\end{compactitem}
An execution $\F$ is said to be \emph{accepting} if and only if the
frontier of each tree $T \in \F$ is labeled with configurations
$q(d_1,\ldots,d_{\#(q)})$ such that $q \in F$ is a final predicate. In
this case, we say that $\A$ accepts $w$ and denote by $\lang{\A}$ the
set of words accepted by $\A$. In this paper, we answer the following
questions: \begin{compactenum}
\item \emph{closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

In the upcoming developments it is more convenient to work with
logical formulae defining executions. Let $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the rest of
this session. For any $i \in \nat$, we denote by $\stamp{Q}{i} =
\set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} = \set{\stamp{x}{i}
  \mid x \in X}$ the sets of time-stamped predicates and variables,
respectively. For a formula $\psi$ and $i \in \nat$, we define
$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
formula in which all input variables and state predicates are replaced
by their time-stamped counterparts. Given a sequence of input events
$\alpha = a_1 \ldots a_n$, the \emph{path formula} of $\alpha$ in $\A$
is defined as follows:
\[\pathform{\A}{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow \stamp{\psi}{i}\]
To formalize the relation between the low-level configuration-based
execution semantics and the symbolic path formulae, consider a word
$w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any execution forest $\F$ of $\A$
over $w$ is associated an interpretation $\I_{\F}$ of the set of
time-stamped predicates $\set{\stamp{q}{i} \mid q \in Q, i \in [n]}$,
defined as follows:
\[\I_{\F}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}}
  \mid q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in
    $\F$}}\enspace.\]

\begin{lemma}\label{lemma:path-formula}
  For any word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have: 
  \[\minsem{\pathform{\A}{w_\Sigma}}_{w_\Data} = \set{\I_{\F} \mid \F
    \text{ is an execution of $\A$ over $w$}}\enspace.\]
\end{lemma}
